Acceleration of plague outbreaks in the second pandemic

### 2020-06-19 ms_text.tex
Main MS; nothing pulled in except by chaining (there should be few or no excess files)
more natural affiliation order

\author{David J.\,D.\ Earn$^{a,c,e}$, 
Junling Ma$^{f}$, 
Hendrik N.\ Poinar$^{b,c,d,e}$,\\
Jonathan Dushoff$^{a,c,e}$,
and Benjamin M.\ Bolker$^{a,c,e}$\\
\href{mailto:earn@math.mcmaster.ca}{\tt earn@math.mcmaster.ca}\\
\small $^a$Department of Mathematics \& Statistics,\\
\small $^b$McMaster Ancient DNA Centre, Department of Anthropology,\\
\small $^c$Department of Biology, $^d$Department of Biochemistry,\\
\small and $^e$Michael G.\ deGroote Institute for Infectious Disease Research,\\
\small McMaster University, 1280 Main Street West, Hamilton, ON, L8S 4K1, Canada;\\
\small $^f$Department of Mathematics \& Statistics, University of Victoria, Victoria, BC, Canada}

@@

%
\newcommand{\pixht}{0.3\textheight}
\begin{figure}
\begin{center}
\mbox{
\hspace{1.2cm}
\includegraphics[height=\pixht]{images/WillExample1644_page86_1.pdf}
%% \includegraphics[height=\pixht]{WillExample1644_page86_1.eps}
\hspace{0.4cm}
\includegraphics[height=\pixht]{images/cripplegate_wellcome_parish.jpg}
%% \includegraphics[height=\pixht]{cripplegate_wellcome_parish.eps}
\hspace{0.4cm}
\includegraphics[height=\pixht]{images/bill1665adjust.pdf}
%% \includegraphics[height=\pixht]{bill1665adjust.eps}
}
\\ \vspace{0.5cm}
\includegraphics[width=\textwidth]{analysis/plots/plot_all.pdf}
%% \includegraphics[width=\textwidth]{plot_all.eps}
\end{center}
\vspace{-0.5cm}
\caption[]{
  \emph{Top left:} Part of a will proved in the Prerogative Court of Canterbury (PCC), dated 18 December 1644 \cite{Archives2018}.
  \emph{Top centre:} A parish register page from St Giles without Cripplegate, August 1665 \cite[courtesy of the Wellcome Collection: \url{https://wellcomecollection.org/works/uutd7scc}]{Bell24}.
  \emph{Top right:} One of the London Bills of Mortality (LBoM), for the week beginning 26 September 1665 (photo by Claire Lees).
\emph{Bottom:}
Mortality in London, England, 1340--1380 and 1540--1680, aggregated 4-weekly, plotted on a log scale.  The three distinct sources of data (see \cref{tab:datasources}) are: 
last wills and testaments of Londoners whose wills were probated in the the Court of Husting (14th c.{}) or the PCC (16th-17th c.{}),
weekly aggregations of burials listed in extant parish registers \cite{Slac90}, 
and weekly plague deaths listed in the LBoM \cite{Crei65}.
Major plague years are highlighted in yellow.  Aside from these and various minor plague years, there are notably unusual patterns during an epidemic of ``sweating sickness'' in 1551 \cite[p.\,70]{Slac90}, the ``influenza'' epidemic of 1557--9 \cite[p.\,70]{Slac90} (which coincided with the end of the reign of ``Bloody'' Mary~I and the ascension of Elizabeth~I in 1558 [indicated by a crown icon]), and the absence of a monarch during the Interregnum from 30 January 1649 to 29 May 1660.
}
\label{F:plot3sources}
\end{figure}
%

@@

\begin{table}[h!] 
  \caption[Estimated $r$ and $1/r$]{Maximum likelihood estimates (MLEs) of the initial exponential growth rate ($r$) and doubling time ($(\log2)/r$) with their 95\% confidence intervals (CIs), obtained from the time series shown in \cref{F:timeseries} (see \Methodslink).  The goodness of fit measure ($R^2$) is the proportional reduction in the mean squared error, \ie $1-d^2/\sigma^2$ where $d^2$ is the model mean squared error and $\sigma^2$ is the (population) variance of the data; predictions and observations are aggregated to weekly before computing the $R^2$ for wills, for consistency among data sources. The implied basic reproduction numbers ($\rzerop$) and attack rates ($\Zp$), assuming the generation interval distribution for Modern pneumonic plague \cite{GaniLeac04}, are also shown.
All MLEs and associated CIs are shown in \cref{F:combdata}.
}
\begin{center}
  \footnotesize
  %%\rowcolors{2}{white}{verylightgray}
  \rowcolors{2}{white}{gray}
  \input{analysis/tables/fakeparams.tex}

@@

The plague mortality data were photographed and/or entered by Seth Earn, Kelly Hancock, Claire Lees, James McDonald, David Price and Hannah Price.  Valerie Hart facilitated work at the Guildhall Library in London.  Josephine Earn provided historical context.  Non-exponential comparison fits were conducted by David Champredon.  Sang Woo Park assisted in finding published estimates of auxiliary epidemic parameters. Alex Bushby assisted with compiling probate dates. Sigal Balshine, Ann Carmichael and David Price made helpful comments.  Early stages of this research were supported by a J.\,S.~McDonnell Foundation Research Award to DE (\url{https://www.jsmf.org/grants/d.php?id=2006014}).  All the authors were supported by Discovery grants from the Natural Sciences and Engineering Research Council of Canada (NSERC). HP was supported by the Social Sciences and Humanities Research Council of Canada (SSHRC). SHARCnet (\url{http://www.sharcnet.ca}) provided computational resources.

>>>>

\author{David J.\,D.\ Earn$^{a,b,c}$, 
Junling Ma$^{f}$, 
Hendrik N.\ Poinar$^{b,c,d,e}$,\\
Jonathan Dushoff$^{a,b,c}$,
and Benjamin M.\ Bolker$^{a,b,c}$\\
\href{mailto:earn@math.mcmaster.ca}{\tt earn@math.mcmaster.ca}\\
\small $^a$Department of Mathematics \& Statistics, $^b$Department of Biology, \\
\small $^c$Michael G.\ deGroote Institute for Infectious Disease Research,\\
\small $^d$McMaster Ancient DNA Centre, Department of Anthropology,\\
\small and $^e$Department of Biochemistry,\\
\small McMaster University, 1280 Main Street West, Hamilton, ON, L8S 4K1, Canada;\\
\small $^f$Department of Mathematics \& Statistics, University of Victoria, Victoria, BC, Canada}

@@

%
\newcommand{\pixht}{0.3\textheight}
\begin{figure}
\begin{center}
\mbox{
\hspace{1.2cm}
\includegraphics[height=\pixht]{images/WillExample1644_page86_1.pdf}
%% \includegraphics[height=\pixht]{WillExample1644_page86_1.eps}
\hspace{0.4cm}
\includegraphics[height=\pixht]{images/cripplegate_wellcome_parish.jpg}
%% \includegraphics[height=\pixht]{cripplegate_wellcome_parish.eps}
\hspace{0.4cm}
\includegraphics[height=\pixht]{images/bill1665adjust.pdf}
%% \includegraphics[height=\pixht]{bill1665adjust.eps}
}
\\ \vspace{0.5cm}
\includegraphics[width=\textwidth]{analysis/plots/plot_allsources.pdf}
%% \includegraphics[width=\textwidth]{plot_allsources.eps}
\end{center}
\vspace{-0.5cm}
\caption[]{
  \emph{Top left:} Part of a will proved in the Prerogative Court of Canterbury (PCC), dated 18 December 1644 \cite{Archives2018}.
  \emph{Top centre:} A parish register page from St Giles without Cripplegate, August 1665 \cite[courtesy of the Wellcome Collection: \url{https://wellcomecollection.org/works/uutd7scc}]{Bell24}.
  \emph{Top right:} One of the London Bills of Mortality (LBoM), for the week beginning 26 September 1665 (photo by Claire Lees).
\emph{Bottom:}
Mortality in London, England, 1340--1380 and 1540--1680, aggregated 4-weekly, plotted on a log scale.  The three distinct sources of data (see \cref{tab:datasources}) are: 
last wills and testaments of Londoners whose wills were probated in the the Court of Husting (14th c.{}) or the PCC (16th-17th c.{}),
weekly aggregations of burials listed in extant parish registers \cite{Slac90}, 
and weekly plague deaths listed in the LBoM \cite{Crei65}.
Major plague years are highlighted in yellow.  Aside from these and various minor plague years, there are notably unusual patterns during an epidemic of ``sweating sickness'' in 1551 \cite[p.\,70]{Slac90}, the ``influenza'' epidemic of 1557--9 \cite[p.\,70]{Slac90} (which coincided with the end of the reign of ``Bloody'' Mary~I and the ascension of Elizabeth~I in 1558 [indicated by a crown icon]), and the absence of a monarch during the Interregnum from 30 January 1649 to 29 May 1660.
}
\label{F:plot3sources}
\end{figure}
%

@@

\begin{table}[h!] 
  \caption[Estimated $r$ and $1/r$]{Maximum likelihood estimates (MLEs) of the initial exponential growth rate ($r$) and doubling time ($(\log2)/r$) with their 95\% confidence intervals (CIs), obtained from the time series shown in \cref{F:timeseries} (see \Methodslink).  The goodness of fit measure ($R^2$) is the proportional reduction in the mean squared error, \ie $1-d^2/\sigma^2$ where $d^2$ is the model mean squared error and $\sigma^2$ is the (population) variance of the data; predictions and observations are aggregated to weekly before computing the $R^2$ for wills, for consistency among data sources. The implied basic reproduction numbers ($\rzerop$) and attack rates ($\Zp$), assuming the generation interval distribution for Modern pneumonic plague \cite{GaniLeac04}, are also shown.
All MLEs and associated CIs are shown in \cref{F:combdata}.
}
\begin{center}
  \footnotesize
  %%\rowcolors{2}{white}{verylightgray}
  \rowcolors{2}{white}{gray}
  \input{analysis/tables/params.tex}

@@

The plague mortality data were photographed and/or entered by Seth Earn, Kelly Hancock, Claire Lees, James McDonald, David Price and Hannah Price.  Valerie Hart facilitated work at the Guildhall Library in London.  Non-exponential comparison fits were conducted by David Champredon.  Sang Woo Park assisted in finding published estimates of auxiliary epidemic parameters. Alex Bushby assisted with compiling probate dates. Sigal Balshine, Ann Carmichael and David Price made helpful comments.  Early stages of this research were supported by a J.\,S.~McDonnell Foundation Research Award to DE (\url{https://www.jsmf.org/grants/d.php?id=2006014}).  All the authors were supported by Discovery grants from the Natural Sciences and Engineering Research Council of Canada (NSERC). HP was supported by the Social Sciences and Humanities Research Council of Canada (SSHRC). SHARCnet (\url{http://www.sharcnet.ca}) provided computational resources.  David Earn dedicates this paper to Josephine Earn, who enthusiastically discussed the history of London (and plagues) with him over many years.
